Let's talk about Thurstone & Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2025-05-14 paper2.qmd
section:5; process:write,equation,code

<!-- 6. error assumptions -->
Next, we specify the probabilistic assumptions for the idiosyncratic errors $e_{I}$, $e_{IA}$, $e_{J}$, and $e_{JK}$. These error terms exhibit indeterminacies in their *location*, *orientation*, and *scale* due to the lack of an inherent scale in the associated latent variables. We must resolve these indeterminacies to identify the model [@Depaoli_2021; @deAyala_2009]. Drawing on principles from SEM [@Hoyle_et_al_2023], we assume that the vector of idiosyncratic errors $\boldsymbol{e} = [e_{I}, e_{IA}, e_{J}, e_{JK}]^{T}$, follows a Multivariate Normal distribution with mean vector $\boldsymbol{\mu}$ and a covariance matrix $\boldsymbol{\Sigma} = \boldsymbol{D} \boldsymbol{R} \boldsymbol{D}$, with $\boldsymbol{D}$ denoting a diagonal matrix of standard deviations and $\boldsymbol{R}$ a correlation matrix. To address the *location* indeterminacy, we set the errors' mean vector to zero: 
$$
\boldsymbol{\mu} = [0, 0, 0, 0]^{T}
$$ {#eq-mat53}

Following [SCM -@fig-cj16_scm], we assume uncorrelated errors to resolve the *orientation* indeterminacy. Thus, we define the correlation matrix $\boldsymbol{R}$ as:
$$
\boldsymbol{R} = \begin{bmatrix}
    1 & 0 & 0 & 0 \\
    0 & 1 & 0 & 0 \\
    0 & 0 & 1 & 0 \\
    0 & 0 & 0 & 1 
  \end{bmatrix}
$$ {#eq-mat54}

@@

Here, $s_{XI}$ represents the standard deviation for the individuals, $p_{IA}$ for the stimuli, $s_{ZJ}$ for the judges, and $p_{JK}$ for the judgments. We assume $s_{XI}$ varies depending on the teaching method group to which each student belongs. In the example from @sec-theoretical, where the teaching method $X_{I} = \{1,2\}$, the model sets the constraint according to Equation [-@eq-mat56]. This constraint anchors the scale of the individuals' latent trait while relaxing the assumption of equal dispersion for the stimuli, thereby addressing the concerns raised in @sec-theory-issue1a. 
$$
\sum_{g=1}^{2} s_{XI}[g]/2 = 1
$$ {#eq-mat56}

@@

Similarly, we assume $s_{ZJ}$ varies depending on the groups to which each judge belongs. For instance, if $Z_{J} = \{1,2,3\}$ represents three groups of judges with varying expertise, the model sets the constraint: 
$$
\sum_{g=1}^{2} s_{ZJ}[g]/3 = 1
$$ {#eq-mat56}

This constraint anchors the scale of the judges' latent trait and relaxes the assumption of equal dispersion for the judgments. Conversely, $p_{JK}$ is defined as a proportion of $1$ to establish the scale of the judgments' latent trait relative to the scale of the judges' trait.

>>>>

<!-- 6. error assumptions -->
Next, we specify the probabilistic assumptions for the idiosyncratic errors $e_{I}$, $e_{IA}$, $e_{J}$, and $e_{JK}$. These error terms exhibit indeterminacies in their *location*, *orientation*, and *scale* due to the lack of an inherent scale in the associated latent variables. Thus, we must resolve these indeterminacies to identify the latent variable model [@Depaoli_2021; @deAyala_2009]. Drawing on principles from SEM [@Hoyle_et_al_2023], we assume that the vector of idiosyncratic errors $\boldsymbol{e} = [e_{I}, e_{IA}, e_{J}, e_{JK}]^{T}$, follows a Multivariate Normal distribution with mean vector $\boldsymbol{\mu}$ and a covariance matrix $\boldsymbol{\Sigma} = \boldsymbol{D} \boldsymbol{R} \boldsymbol{D}$, with $\boldsymbol{D}$ denoting a diagonal matrix of standard deviations and $\boldsymbol{R}$ a correlation matrix. To address the *location* indeterminacy, we set the errors' mean vector to zero: 
$$
\boldsymbol{\mu} = [0, 0, 0, 0]^{T}
$$ {#eq-mat53}

Following [SCM -@fig-cj16_scm], we resolve the *orientation* indeterminacy by assuming that the errors are uncorrelated. This assumption leads us to specify a correlation matrix for the errors, $\boldsymbol{R}$, as:
$$
\boldsymbol{R} = \begin{bmatrix}
    1 & 0 & 0 & 0 \\
    0 & 1 & 0 & 0 \\
    0 & 0 & 1 & 0 \\
    0 & 0 & 0 & 1 
  \end{bmatrix}
$$ {#eq-mat54}

@@

Here, $s_{XI}$ represents the standard deviation for the individuals, $p_{IA}$ for the stimuli, $s_{ZJ}$ for the judges, and $p_{JK}$ for the judgments. We assume $s_{XI}$ varies depending on the teaching method group to which each student belongs. In the example from @sec-theoretical, where the teaching method $X_{I} = \{1,2\}$, the model sets the constraint according to Equation ([-@eq-mat56]). This constraint anchors the scale of the individuals' latent trait while relaxing the assumption of equal dispersion for the stimuli, thereby addressing the concerns raised in @sec-theory-issue1a. 
$$
\sum_{g=1}^{2} s_{XI}[g]/2 = 1
$$ {#eq-mat56}

@@

Similarly, we assume $s_{ZJ}$ varies depending on the groups to which each judge belongs. For instance, if $Z_{J} = \{1,2,3\}$ represents three groups of judges with varying expertise, the model sets the constraint according to Equation ([-@eq-mat57]). This constraint anchors the scale of the judges' latent trait and relaxes the assumption of equal dispersion for the judgments. 
$$
\sum_{g=1}^{2} s_{ZJ}[g]/3 = 1
$$ {#eq-mat57}

Conversely, $p_{JK}$ is defined as a proportion of $1$ to establish the scale of the judgments' latent trait relative to the scale of the judges' trait.
